---
title: >
  Type I Error and Power of Exact and Asymptotic Trend
  Tests in Small Ordered $r \times 2$ Tables: A
  Simulation Study
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
linestretch: 1.1
geometry: 'left=3cm,right=3cm,top=2.5cm,bottom=2.5cm'
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage[mathlines]{lineno}
  - \linenumbers
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4,
  cache = TRUE
)
library(knitr)
library(kableExtra)
```

# Abstract {-}

Trend tests for ordered $r \times 2$ tables are usually
applied either as a large-sample Cochran-Armitage
statistic or as one of several existing exact conditional
implementations; in small, sparse samples the choice
between them is often left unexamined. We report a small
ADEMP-structured Monte Carlo comparison of three trend
procedures -- the asymptotic Cochran-Armitage test, the
unordered Fisher exact test, and the exact conditional
Cochran-Armitage test implemented in the `CATTexact`
package -- under a null and a dose-response alternative
with $k = 4$ groups of size 10 ($B = 2{,}000$
replications per scenario). We find that the realized
type I error rates of the three tests differ
substantially at nominal $\alpha = 0.05$, and we identify
the doubled one-sided `CATTexact` p-value used for the
two-sided exact test as a conservative operational choice
that plausibly explains part of this spread; we discuss
its effect on the power comparison and report the
observed rejection rates directly rather than only their
expected direction. We do not implement or benchmark the
tree-based pruning algorithm sketched in Section
\@ref(pruning-adaptations) as a proposed extension; the
contribution of this report is the simulation comparison
itself, not new enumeration software.

# Introduction

## The problem of testing for trend

Clinical trials and observational studies frequently
compare binomial proportions across ordered groups. A
dose-response study, for example, may assign subjects to
$k$ increasing dose levels. Each subject's binary outcome
is then recorded. The scientific question is not merely
whether the groups differ. That is the omnibus hypothesis
tested by Pearson's $\chi^2$ or Fisher's exact test. We
are instead interested in whether the proportions exhibit
a monotone trend consistent with the ordering of the
groups. This distinction matters. An omnibus test that
ignores the ordering sacrifices power against the ordered
alternative that is typically of primary interest
[@cochran1954methods; @armitage1955tests].

## The Cochran-Armitage trend test

@cochran1954methods proposed a partitioning of the
Pearson $\chi^2$ statistic into components corresponding
to polynomial contrasts. He noted that the linear
component provides a powerful test against a monotone
trend. @armitage1955tests formalized this approach for
$2 \times k$ tables. He defined the trend test statistic
\[
T = \sum_{i=1}^{k} d_i \, X_i
  - \frac{R_1 \sum_{i=1}^{k} d_i \, n_i}{N},
\]
Here $X_i$ is the number of events in group $i$, $n_i$
is the group size, $R_1 = \sum X_i$ is the total event
count, and $N = \sum n_i$ is the grand total. The scores
$d_1 < d_2 < \cdots < d_k$ are assigned to the ordered
groups. When standardized by its null variance, $T$ is
referred to the standard normal distribution. This is
the Cochran-Armitage trend test. It is now standard in
regulatory submissions and is widely implemented in
statistical software
[@liu2005analysis; @hothorn2008implementing].

## Limitations of the asymptotic approximation

The normal approximation to $T$ requires moderately large
samples. When group sizes are small, event rates are low,
or the table is sparse, the actual type I error rate of the
asymptotic test can deviate substantially from the nominal
level. @williams1988tests demonstrated this problem for
balanced designs. He proposed an exact conditional
alternative. @lin2006exact compared exact and asymptotic
versions under several models. They confirmed that the
asymptotic test can be anticonservative in small samples.
These concerns are essentially analogous to those that
motivate Fisher's exact test as a replacement for
Pearson's $\chi^2$ in $2 \times 2$ tables. The same
concerns extend, by analogy, to general $r \times c$
tables.

## Exact approaches to the trend test

The exact conditional approach conditions on the observed
marginal totals. It then enumerates, or efficiently
computes, the null distribution of the trend statistic
over the reference set of all tables with those margins.
This strategy was pioneered by @patefield1982exact. He
recommended exact tests based on the $\chi^2$ trend
statistic for ordered contingency tables. @mehta1984exact
developed exact significance testing for ordered
categorical data using the network algorithm of
@mehta1983network. @agresti1990exact extended exact
conditional inference to contingency tables with ordered
categories. He proposed an efficient numerical algorithm
that exploits the linear-by-linear association structure.

The computational machinery for these exact tests was
further developed by @mehta1992exact. He presented an
algorithm for exact stratified linear rank tests that
encompasses the exact Cochran-Armitage test as a special
case. @agresti1992survey provided a comprehensive review
of exact inference for contingency tables. That review
situates trend tests within the broader conditional
inference framework. @agresti2001exact updated this
review with attention to continuing controversies. Among
these is the debate between conditional and unconditional
exact tests.

## Conditional versus unconditional exact tests

Two distinct approaches to exact inference exist. The
conditional approach fixes both sets of marginal totals.
It then enumerates tables from the multivariate
hypergeometric distribution. The unconditional approach
fixes only the group sizes, that is, the column totals.
It maximizes or integrates over the nuisance parameter,
the common event probability under $H_0$.
@shan2012efficient developed an efficient unconditional
exact test for trend using Lloyd's approach. That work
demonstrated superior power over the conditional test in
many configurations. @consiglio2014comparison compared
conditional and unconditional exact trend tests using
Bartholomew's statistic. They noted that the conditional
test is somewhat conservative but guarantees type I
error control [@bartholomew1959test].

## Current standard implementations

The dominant computational approach to the exact
conditional trend test is the network algorithm of
@mehta1980network and @mehta1992exact. This algorithm
reformulates the enumeration of tables with fixed margins
as a longest-path problem in a directed acyclic network.
It underlies the commercial software StatXact
[@mehta1983network]. It is also understood to underlie
exact trend testing in SAS PROC FREQ and SPSS Exact
Tests, though we have not verified this against the
SAS/SPSS documentation directly.

In R, three packages provide exact or near-exact trend
tests. The `CATTexact` package [@edelmann2025catexact]
implements the conditional exact Cochran-Armitage test
using the @mehta1992exact algorithm. It computes the
one-sided exact p-value via network enumeration. The
`coin` package [@hothorn2006lego; @hothorn2008implementing]
provides a general permutation test framework based on
the theory of @mehta1992exact and
@strasser1999asymptotic. The trend test is obtained there
as a special case of `independence_test()` with ordered
factors, using either exact enumeration, Monte Carlo
approximation, or the asymptotic distribution. The `perm`
package offers exact permutation trend tests via both
network and complete enumeration algorithms. The latter,
however, is limited to small total sample sizes.

Despite this software ecosystem, all existing exact
implementations rely on the Mehta-Patel network
algorithm or on direct permutation enumeration. We note
that no implementation exploits tree-based traversal with
memoized pruning. That approach has proven effective for
the unordered Fisher exact test in $r \times 2$ tables.
The network algorithm enumerates over a $(k, c_1)$ state
space. It aggregates contributions from all tables that
share the same partial column sum at each stage. The
tree-based approach, by contrast, traverses the space of
individual cell assignments. It prunes entire subtrees
using bounds on the test statistic. For the
probability-ordered Fisher test, we have shown that
tree-based pruning can be competitive with, or faster
than, the network algorithm on small to moderate tables.
This advantage is most apparent when the table is sparse
or has unequal row margins. Whether these advantages
transfer to the statistic-ordered trend test is an open
question. The present study begins to address it.

The unconditional exact approach fixes only group sizes
and maximizes over the nuisance parameter. It has been
developed by @shan2012efficient and
@consiglio2014comparison, but remains less widely
adopted. It offers better power in many configurations.
On the other hand, it is computationally more expensive
and lacks the elegant connection to the multivariate
hypergeometric distribution that makes the conditional
approach algorithmically tractable. @hirji2006exact
provides a comprehensive treatment of both conditional
and unconditional exact methods for discrete data.

## The present study

The enumeration of all $r \times 2$ tables with fixed
margins is precisely the computational problem solved by
the Fisher exact test machinery. In a companion project,
we developed efficient tree-based and network-based
algorithms for computing Fisher's exact p-value in
general $r \times 2$ tables. These use memoized pruning
strategies that achieve substantial speedups over
existing implementations. It is a natural conjecture that
this same enumeration machinery could, in principle, be
adapted to the Cochran-Armitage trend statistic. This
would require replacing the probability-based acceptance
criterion with a statistic-based one. Section
\@ref(pruning-adaptations) sketches the pruning bounds
such an adaptation would require. We have not, however,
implemented or benchmarked this adaptation in the
present work. It is recorded instead as a proposed
extension (Section \@ref(future-research), item 1), not
as a completed contribution.

What we report here instead is a small, ADEMP-structured
Monte Carlo comparison of the trend tests already
available for small, sparse ordered tables. We pursue two
objectives. The first is to compare the type I error and
power of the asymptotic Cochran-Armitage test, the
unordered Fisher exact test, and the exact conditional
Cochran-Armitage test as implemented in the existing
`CATTexact` package, under a null and a dose-response
alternative. The second is to report these operating
characteristics honestly. This includes an unresolved
discrepancy between the exact test's realized size and
that of the unordered Fisher test, which we discuss but
do not fully explain (Section \@ref(discussion)).

# Methods

## The exact conditional trend test

Consider a $k \times 2$ contingency table with row
margins $n_1, \ldots, n_k$ (group sizes) and column
margins $R_1, R_2$ (total events and nonevents). Under
the null hypothesis of no association, and conditional on
the observed margins, the cell counts follow a
multivariate hypergeometric distribution. Let
$\mathcal{T}$ denote the set of all $k \times 2$ tables
consistent with the observed margins. For each table
$t \in \mathcal{T}$, we define the trend statistic
\[
T(t) = \sum_{i=1}^{k} d_i \, x_i(t),
\]
Here $x_i(t)$ is the first-column entry in row $i$ of
table $t$. The two-sided exact p-value is
\[
p = \sum_{t \in \mathcal{T}}
    \Pr(t) \cdot
    \mathbf{1}\!\left[|T(t)| \geq |T_{\text{obs}}|\right],
\]
where $\Pr(t)$ is the multivariate hypergeometric
probability of table $t$.

**Operational two-sided p-value used in this study.** We
note that the sum-based definition above is not
implemented by the `CATTexact` package used for the
simulation in Section \@ref(simulation-design). Instead,
`CATTexact::catt_exact()` returns only a one-sided exact
p-value, computed via the Mehta-Patel network algorithm
[@mehta1992exact; @edelmann2025catexact]. The two-sided
p-values reported in this study are instead obtained by
doubling that one-sided p-value,
$p_{\text{two-sided}} = \min(2 p_{\text{one-sided}}, 1)$.
This is a standard heuristic for constructing a two-sided
p-value from a one-sided exact test when the null
distribution of the test statistic is not guaranteed to
be symmetric [@agresti1992survey]. It is, however, a
distinct construction from the sum-based definition
above. For a discrete, generally asymmetric null
distribution of $T$, doubling is known to be
conservative. It can overstate the two-sided p-value
relative to the mass-based definition. We did not
implement the mass-based two-sided p-value directly.
Doing so would require our own enumeration of
$\mathcal{T}$, which is the unimplemented extension
discussed in Section
\@ref(connection-to-the-fisher-exact-test-framework-proposed).
The doubling rule should therefore be understood as the
operational definition of 'two-sided exact trend p-value'
throughout the Results and Discussion. It is not a
verified implementation of the definition given above.
We return to the practical consequences of this choice
in Section \@ref(discussion), in particular its likely
contribution to the exact trend test's conservatism
relative to the unordered Fisher test under the null.

## Connection to the Fisher exact test framework (proposed)

This subsection and Section \@ref(pruning-adaptations)
describe a proposed adaptation of the companion Fisher
exact test enumeration machinery to the trend statistic.
We note that neither the tree-based traversal nor the
pruning bounds below have been implemented, run, or
benchmarked in this repository. The simulation study in
Section \@ref(simulation-design) uses the existing
`CATTexact` implementation instead (see Section
\@ref(future-research), item 1). The bounds are recorded
here as a mathematically well-defined starting point for
future implementation work. They are not a completed or
evaluated contribution of this report.

In the standard Fisher exact test for $r \times 2$
tables, the p-value is
\[
p_{\text{Fisher}} = \sum_{t \in \mathcal{T}}
    \Pr(t) \cdot
    \mathbf{1}\!\left[\Pr(t) \leq \Pr(t_{\text{obs}})
    \right].
\]
The only difference between the two tests is the
ordering criterion. Fisher's test orders tables by their
probability. The exact trend test, on the other hand,
orders them by the absolute value of the trend statistic.
The enumeration of $\mathcal{T}$ and the computation of
$\Pr(t)$ are identical between the two tests. This means
the tree-based traversal with find\_max/find\_min
pruning, and the network-based state-space enumeration
from our $r \times 2$ Fisher exact test package, can in
principle be adapted to the trend test. The adaptation
would replace the probability-based acceptance criterion
with a statistic-based one.

## Pruning adaptations

These bounds are derived analytically as a proposed
extension (see the note at the start of Section
\@ref(connection-to-the-fisher-exact-test-framework-proposed)).
They have not been coded or tested against a reference
implementation. We make no runtime or correctness claim
for them here.

The key algorithmic question is whether the pruning
strategies developed for the probability ordering
transfer to the statistic ordering. In the
probability-ordered Fisher test, the find\_max and
find\_min bounds allow entire subtrees to be skipped.
This occurs when all tables in the subtree have
probability above or below the threshold. For the trend
test, analogous bounds on the trend statistic are
needed.

Consider a partial table with values $x_1, \ldots, x_j$
assigned to the first $j$ rows, and remaining column
margin $c_1' = R_1 - \sum_{i=1}^{j} x_i$. The minimum
and maximum achievable values of $T$ over all completions
are
\begin{align*}
T_{\min} &= \sum_{i=1}^{j} d_i x_i +
  \min_{\{x_{j+1},\ldots,x_k\}}
  \sum_{i=j+1}^{k} d_i x_i, \\
T_{\max} &= \sum_{i=1}^{j} d_i x_i +
  \max_{\{x_{j+1},\ldots,x_k\}}
  \sum_{i=j+1}^{k} d_i x_i,
\end{align*}
where the optimization is subject to the margin
constraints. These bounds can be computed in $O(k-j)$
time by a greedy allocation of the remaining column
margin to the rows with the largest, or smallest,
scores. If $|T_{\max}| < |T_{\text{obs}}|$ and
$|T_{\min}| < |T_{\text{obs}}|$, the entire subtree can
be pruned.

## Simulation design

```{r sim-setup}
source("../scripts/sim_trend_test.R")

# Morris, White & Crowther (2019) §4.1: set RNGkind and seed once at
# program start. Both scenarios below draw from the same L'Ecuyer stream
# in sequence so that any difference between null and alternative is
# attributable to the DGM, not to the RNG state.
RNGkind("L'Ecuyer-CMRG")
set.seed(20260417)
```

### ADEMP structure

This simulation study is reported following the ADEMP framework of
Morris, White, and Crowther [-@morris2019using].

**Aims.** Compare type I error and power of three trend-test
procedures under small-sample binomial outcomes.

**Data-generating mechanisms.** Two scenarios, both with $k = 4$
groups of size $n_i = 10$ and scores $d = (1, 2, 3, 4)$:

- Null: $p_i = 0.2$ for all groups.
- Alternative: $p = (0.05, 0.10, 0.20, 0.35)$ representing a
  dose-response relationship.

**Estimands.** The binary decision to reject $H_0: \text{no trend}$
at $\alpha = 0.05$.

**Methods.** (i) Asymptotic Cochran-Armitage trend test (two-sided);
(ii) Fisher's exact test (unordered); (iii) Exact conditional
Cochran-Armitage trend test via `CATTexact::catt_exact()` with the
one-sided p-value doubled for two-sided comparison.

**Performance measures.** Rejection rate at $\alpha = 0.05$, mean
p-value, and median p-value. Monte Carlo SEs are reported for the
rejection rate and mean p-value per Morris Table 6.

**Simulation size.** $B = 2{,}000$ replications per scenario. For the
null scenario at $p \approx 0.05$, the Monte Carlo SE for rejection
rate is $\sqrt{0.05 \times 0.95 / 2000} \approx 0.005$. For the
alternative at $p \approx 0.8$, the Monte Carlo SE is
$\sqrt{0.8 \times 0.2 / 2000} \approx 0.009$. Both are well below
$1$ pp. $B$ was chosen as a round number and verified post hoc
against these targets. It was not derived a priori from a
worst-case ($p = 0.5$) target MCSE, which would have required
$B \geq 2{,}500$ (see Limitations, Section \@ref(limitations)).

### Morris et al. (2019) ADEMP compliance

We audited this simulation study against the reporting
standards of Morris, White, and Crowther
[-@morris2019using]. An initial audit
(`docs/morris-audit-2026-04-17.md`, dated 2026-04-17)
found the study non-compliant. That audit cited three
problems: (a) `set.seed()` was called inside
`sim_trend_test()` and reseeded on every call; (b) two
distinct per-scenario seeds prevented a paired
cross-scenario comparison; and (c) `summarize_sim()`
reported no Monte Carlo SE. Those issues have since been
remediated in the current code. `sim_trend_test()` no
longer calls `set.seed()` (see the comment at
`sim_trend_test.R:6-7`). A single
`RNGkind("L'Ecuyer-CMRG")` and `set.seed(20260417)` are
now set once above, before both scenarios run in
sequence from the same stream. `.Random.seed` is
captured per replication (`sim_trend_test.R:17`). And
`summarize_sim()` now reports `mcse_reject` and
`mcse_mean_p` for every method (`sim_trend_test.R:88,
90`), shown in Tables 1-2. As of this revision, the
study is compliant with the seed-discipline and
MCSE-reporting criteria audited above. One residual gap
remains: $B = 2{,}000$ was verified post hoc rather than
derived a priori from a target MCSE. This is noted in
the Simulation size paragraph above and discussed
further in Limitations (Section \@ref(limitations)).

```{r sim-null, cache=TRUE}
sim_null <- sim_trend_test(
  n_per_group = rep(10, 4),
  true_probs = rep(0.2, 4),
  scores = 1:4,
  n_rep = 2000
)
sim_null$p_exact_trend <- pmin(
  2 * sim_null$p_exact_trend, 1
)
```

```{r sim-alt, cache=TRUE}
sim_alt <- sim_trend_test(
  n_per_group = rep(10, 4),
  true_probs = c(0.05, 0.10, 0.20, 0.35),
  scores = 1:4,
  n_rep = 2000
)
sim_alt$p_exact_trend <- pmin(
  2 * sim_alt$p_exact_trend, 1
)
```

# Results

## Type I error control (Scenario 1)

```{r table-null}
tab_null <- summarize_sim(sim_null)

fmt_rate <- function(tab, meth) {
  sprintf("%.3f", tab$reject_rate[tab$method == meth])
}
r_null_asymp <- fmt_rate(tab_null, "CA Asymptotic")
r_null_fisher <- fmt_rate(tab_null, "Fisher Exact (unordered)")
r_null_exact <- fmt_rate(tab_null, "CA Exact (conditional)")

kable(
  tab_null,
  format = "latex",
  booktabs = TRUE,
  digits = 3,
  col.names = c(
    "Method", "N valid", "Reject rate",
    "MCSE reject", "Mean p", "MCSE mean p",
    "Median p"
  ),
  caption = paste(
    "Type I error rates under the null hypothesis.",
    "Four groups, $n_i = 10$, $p_i = 0.2$,",
    "$\\alpha = 0.05$, $B = 2{,}000$."
  ),
  escape = FALSE
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

Table 1 presents the rejection rates under the null
hypothesis. The asymptotic Cochran-Armitage test rejected
at `r r_null_asymp`, close to the nominal 0.05. The
unordered Fisher exact test rejected at `r r_null_fisher`.
The exact conditional trend test rejected at
`r r_null_exact`, markedly below both the nominal level
and the unordered Fisher test's realized size. This
ordering is the opposite of what conditioning on a
strict refinement of the reference set would suggest, if
size alone were at issue. We attribute it at least in
part to the doubling-based construction of the exact
trend test's two-sided p-value, described in Section
\@ref(the-exact-conditional-trend-test). We discuss this
anomaly further in Section \@ref(discussion).

## Power comparison (Scenario 2)

```{r table-alt}
tab_alt <- summarize_sim(sim_alt)
r_alt_asymp <- fmt_rate(tab_alt, "CA Asymptotic")
r_alt_fisher <- fmt_rate(tab_alt, "Fisher Exact (unordered)")
r_alt_exact <- fmt_rate(tab_alt, "CA Exact (conditional)")

n_dropped_alt <- 2000L - min(tab_alt$n_valid)
kable(
  tab_alt,
  format = "latex",
  booktabs = TRUE,
  digits = 3,
  col.names = c(
    "Method", "N valid", "Reject rate",
    "MCSE reject", "Mean p", "MCSE mean p",
    "Median p"
  ),
  caption = paste(
    "Power under the alternative hypothesis.",
    "Four groups, $n_i = 10$,",
    "$p = (0.05, 0.10, 0.20, 0.35)$,",
    "$\\alpha = 0.05$, $B = 2{,}000$.",
    "N valid is below $B$ for the asymptotic and exact",
    "trend methods because",
    n_dropped_alt,
    "replications produced a degenerate column margin",
    "($c_1 = 0$ or $c_1 = N$), for which the",
    "Cochran-Armitage variance is zero and both methods",
    "return \\texttt{NA}; see the text for detail."
  ),
  escape = FALSE
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

Table 2 shows the power of each method to detect the
dose-response trend. The asymptotic Cochran-Armitage
test rejected at `r r_alt_asymp`. The unordered Fisher
exact test rejected at `r r_alt_fisher`, and the exact
conditional trend test rejected at `r r_alt_exact`. Under
the alternative, `r n_dropped_alt` of the 2,000
replications for the asymptotic and exact trend methods
were excluded from the denominator. This occurred
because a degenerate column margin ($c_1 = 0$ or
$c_1 = N$, that is, all groups had the same outcome)
made the Cochran-Armitage variance zero
(`sim_trend_test.R`, `var_num <= 0` branch). The Fisher
exact test does not require this variance, and so
retains all 2,000 replications. The same two replications
are dropped for both affected methods. This is
consistent with a shared cause, the same degenerate
simulated table, rather than two independent failure
modes. Excluding two replications out of 2,000 changes
the denominator by 0.1 percent. We do not expect this to
materially bias the reported rates, but we disclose it
here rather than leave it silent.

At face value, the exact conditional trend test
outperforms the unordered Fisher test
(`r r_alt_exact` vs. `r r_alt_fisher`). This is
consistent with the expectation that exploiting the
ordering increases power. This comparison is, however,
confounded by the unequal realized type I error
documented in Section
\@ref(type-i-error-control-scenario-1). At nominal
$\alpha = 0.05$, the three methods' realized null
rejection rates differ by roughly three-fold
(`r r_null_asymp`, `r r_null_fisher`, `r r_null_exact`
for the asymptotic, Fisher, and exact trend tests,
respectively). Part of the power ranking therefore
reflects differences in conservatism, rather than
differences in statistical efficiency at a common
actual size. We do not report size-adjusted, that is
empirical-alpha-matched, power curves in this study.
Doing so correctly would require re-deriving each
method's rejection threshold from its own null
distribution, which is beyond the scope of the present
simulation. The power figures above should therefore be
read as raw operating characteristics at the nominal,
not the realized, $\alpha$. The power ranking should not
be interpreted as a size-adjusted efficiency ranking. See
Section \@ref(discussion) for further interpretation.

## Distribution of p-values

```{r fig-null, fig.cap="Distribution of p-values under the null hypothesis. Dashed line at $\\alpha = 0.05$.", fig.height=3.5}
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
hist(
  sim_null$p_asymp,
  breaks = 20,
  main = "CA Asymptotic",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
hist(
  sim_null$p_fisher,
  breaks = 20,
  main = "Fisher Exact",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
hist(
  sim_null$p_exact_trend,
  breaks = 20,
  main = "CA Exact",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
```

```{r fig-power, fig.cap="Distribution of p-values under the alternative hypothesis. Dashed line at $\\alpha = 0.05$.", fig.height=3.5}
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
hist(
  sim_alt$p_asymp,
  breaks = 20,
  main = "CA Asymptotic",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
hist(
  sim_alt$p_fisher,
  breaks = 20,
  main = "Fisher Exact",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
hist(
  sim_alt$p_exact_trend,
  breaks = 20,
  main = "CA Exact",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
```

## Agreement between exact tests

```{r fig-agreement, fig.cap="Scatterplot of Fisher exact p-values versus exact Cochran-Armitage p-values under the alternative. Dashed line is the identity.", fig.height=4}
par(mar = c(4, 4, 2, 1))
plot(
  sim_alt$p_fisher,
  sim_alt$p_exact_trend,
  pch = 1,
  cex = 0.5,
  col = "grey40",
  xlab = "Fisher exact p-value",
  ylab = "CA exact p-value",
  main = "Agreement under alternative"
)
abline(0, 1, lty = 2, lwd = 1.5)
```

# Discussion

The simulation results characterize the observed operating
characteristics of the exact conditional Cochran-Armitage
trend test relative to the asymptotic version and the
unordered Fisher exact test, in one small, sparse
data-generating regime. Several findings merit discussion.

First, all three tests controlled or under-controlled
type I error at the nominal level in this scenario
(`r r_null_asymp`, `r r_null_fisher`, and
`r r_null_exact` for the asymptotic, Fisher, and exact
trend tests, respectively; Section
\@ref(type-i-error-control-scenario-1)). The exact trend
test's realized size, however, was substantially below
both the nominal level and the unordered Fisher exact
test's realized size. This is something of an anomaly.
Conditioning on the trend statistic's reference set, a
refinement of the Fisher reference set, would on size
grounds alone be expected to bring the realized size
closer to nominal, not further from it. We attribute
this, at least in part, to the two-sided p-value
construction used for the exact trend test. As described
in Section \@ref(the-exact-conditional-trend-test), the
reported two-sided exact trend p-value is obtained by
doubling `CATTexact`'s one-sided exact p-value. This is
a standard but conservative heuristic for a discrete,
possibly asymmetric null distribution. Doubling can push
many one-sided p-values that sit just above $\alpha / 2$
to just above $\alpha$ once doubled. This discards
rejections that a mass-based two-sided definition might
retain, that is, one summing hypergeometric probability
over $|T(t)| \geq |T_{\text{obs}}|$ (Section
\@ref(the-exact-conditional-trend-test)). We did not
implement the mass-based definition to confirm this
mechanism quantitatively in the present study. Doing so
requires the enumeration machinery discussed in Section
\@ref(connection-to-the-fisher-exact-test-framework-proposed),
which remains a proposed extension. The doubling
construction is consistent with, but not proven to fully
account for, the magnitude of the anomaly. We regard it
as a plausible explanation, not a demonstrated one.

Second, the exact trend test's raw rejection rate under
the alternative (`r r_alt_exact`) exceeded the unordered
Fisher test's (`r r_alt_fisher`). This is consistent with
the expectation that exploiting the ordering increases
power against a monotone alternative. As discussed in
Section \@ref(power-comparison-scenario-2), this
comparison is confounded by the unequal realized sizes
documented above. It should not be read as a size-adjusted
efficiency comparison. This power gain is nonetheless
directionally consistent with the motivation for trend
tests dating to @cochran1954methods and
@armitage1955tests.

Third, this study does not implement or benchmark the
tree-based pruning adaptation sketched in Section
\@ref(pruning-adaptations). The bounds $T_{\min}$ and
$T_{\max}$ described there are a mathematically
well-defined starting point. The enumeration machinery
from the companion Fisher exact test project could
plausibly be adapted to compute exact trend p-values
directly, which would avoid the doubling-rule
conservatism discussed above. This remains future work,
however (Section \@ref(future-research), item 1). No
runtime, correctness, or agreement comparison against the
network algorithm or `CATTexact` has been performed.

The conditional approach has known limitations beyond
the doubling-rule issue discussed above. By conditioning
on both sets of margins, it can be conservative relative
to unconditional exact tests
[@shan2012efficient; @consiglio2014comparison]. This
conservatism arises because the conditional reference
set is a subset of the unconditional reference set,
which can lead to discrete jumps in the p-value
function. Whether this conservatism is acceptable
depends on the regulatory context and on the desired
balance between type I error control and power
[@agresti2001exact].

## Limitations

This study has several limitations. (1) The comparison
covers a single pair of scenarios ($k = 4$ groups of size
10, one null and one alternative configuration). The
relative ranking of the three tests may not generalize to
other group counts, unequal group sizes, unequal score
spacing, or other effect sizes. A referee reviewing this
work for a broader claim would reasonably ask for
additional data-generating mechanisms (Section
\@ref(future-research)). (2) The two-sided exact trend
p-value is constructed by doubling a one-sided exact
p-value, rather than by direct enumeration of the
mass-based two-sided definition in Section
\@ref(the-exact-conditional-trend-test). This operational
choice is disclosed above but not resolved. (3) The
unconditional exact trend test of @shan2012efficient and
the Bartholomew-statistic comparison of
@consiglio2014comparison, both cited as competitors with
claimed higher power than the conditional exact test, are
not included in the empirical comparison. (4) The
tree-based pruning algorithm that motivated this study's
original framing is not implemented or evaluated here
(Section \@ref(pruning-adaptations)). (5) $B = 2{,}000$
replications per scenario was not derived a priori from a
target Monte Carlo SE under the Morris et al. (2019)
framework. It was instead chosen as a round number and
verified post hoc to give MCSE well below 1 percentage
point for the realized rejection proportions in this
study (Section \@ref(simulation-design)). A worst-case
($p = 0.5$) target MCSE derivation, as recommended by
[-@morris2019using], would have required $B \geq 2{,}500$
and was not performed.

## Data and code availability

The simulation driver (`sim_trend_test.R`), the summary
function (`summarize_sim()`), and this manuscript source
are all contained in this repository
(`32-test-for-trend/testtrend`). The simulation depends
on the `CATTexact` package, with its version pinned in
`renv.lock`, for the exact conditional p-value. No other
non-CRAN dependencies are required. The companion
$r \times 2$ Fisher exact test enumeration project
referenced in Sections \@ref(the-present-study) and
\@ref(pruning-adaptations) is a separate repository
(`01-fisher-exact-rx2`), not included here. We note that
the pruning adaptation described in this manuscript has
not been implemented in that project either, and no code
for it currently exists anywhere. This manuscript does
not use any external or restricted data. All results are
generated by Monte Carlo simulation from the
data-generating mechanisms in Section
\@ref(simulation-design), with the RNG seed fixed once
(`set.seed(20260417)` after `RNGkind("L'Ecuyer-CMRG")`)
before both scenarios run in sequence.

# Future research

Several extensions of the current work merit
consideration.

1. **Implementation in C/C++.** The present simulation
   uses the `CATTexact` package for the exact
   conditional test, including its one-sided p-value
   doubled for the two-sided comparison (Section
   \@ref(the-exact-conditional-trend-test)). A direct
   implementation, using the tree-based and network-based
   algorithms from the companion $r \times 2$ Fisher
   exact test project and adapted per Sections
   \@ref(connection-to-the-fisher-exact-test-framework-proposed)
   and \@ref(pruning-adaptations), would enable both
   performance comparisons and a direct, mass-based
   two-sided p-value. This would avoid the doubling-rule
   conservatism discussed in Section \@ref(discussion).
   It would also let the method handle larger tables than
   are currently feasible.

2. **Unconditional exact trend tests.** Following
   @shan2012efficient, the unconditional exact approach
   merits implementation using the same enumeration
   infrastructure. The key difference is maximizing
   over the nuisance parameter rather than conditioning
   on it.

3. **Alternative score functions.** The choice of scores
   $d_i$ affects power. Equally-spaced scores, dose-based
   scores, and data-driven scores have been compared in
   the literature [@lin2006exact]. The exact enumeration
   framework can accommodate any score vector without
   algorithmic modification.

4. **Stratified exact trend tests.** The algorithm of
   @mehta1992exact handles stratification. Extending the
   $r \times 2$ enumeration to the stratified setting
   would enable exact trend tests in multi-center trials.

5. **Order-restricted inference.** @agresti1998order
   developed methods for monotone trend alternatives
   using order-restricted maximum likelihood estimation.
   Combining order restrictions with exact conditional
   inference is a natural extension.

6. **Software package.** A dedicated R package that
   unifies the Fisher exact test and exact trend test
   for $r \times 2$ tables, built on shared C/C++
   enumeration code, would provide a practical tool for
   applied statisticians working with small-sample
   ordered categorical data.

# References
