---
title: >
  An Exact Conditional Test for Trend in $r \times 2$
  Tables via Fisher Enumeration
author: "R. G. Thomas"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4,
  cache = TRUE
)
library(knitr)
library(kableExtra)
```

# Introduction

## The problem of testing for trend

Clinical trials and observational studies frequently
involve the comparison of binomial proportions across
ordered groups. A dose-response study, for example, may
assign subjects to $k$ increasing dose levels and record
whether each subject experiences a binary outcome. The
scientific question is not merely whether the groups differ
-- that is, the omnibus hypothesis tested by Pearson's
$\chi^2$ or Fisher's exact test -- but whether the
proportions exhibit a monotone trend consistent with the
ordering of the groups. The distinction matters: an omnibus
test that ignores the ordering sacrifices power against the
ordered alternative that is typically of primary interest
[@cochran1954methods; @armitage1955tests].

## The Cochran-Armitage trend test

@cochran1954methods proposed a partitioning of the
Pearson $\chi^2$ statistic into components corresponding
to polynomial contrasts, and noted that the linear
component provides a powerful test against a monotone
trend. @armitage1955tests formalized this approach for
$2 \times k$ tables, defining the trend test statistic
\[
T = \sum_{i=1}^{k} d_i \, X_i
  - \frac{R_1 \sum_{i=1}^{k} d_i \, n_i}{N},
\]
where $X_i$ is the number of events in group $i$,
$n_i$ is the group size, $R_1 = \sum X_i$ is the total
event count, $N = \sum n_i$ is the grand total, and
$d_1 < d_2 < \cdots < d_k$ are numerical scores assigned
to the ordered groups. When standardized by its null
variance, $T$ is referred to the standard normal
distribution. This is the Cochran-Armitage trend test,
now standard in regulatory submissions and widely
implemented in statistical software
[@liu2005analysis; @hothorn2008implementing].

## Limitations of the asymptotic approximation

The normal approximation to $T$ requires moderately large
samples. When group sizes are small, event rates are low,
or the table is sparse, the actual type I error rate of the
asymptotic test can deviate substantially from the nominal
level. @williams1988tests demonstrated this problem for
balanced designs and proposed an exact conditional
alternative. @lin2006exact compared exact and asymptotic
versions under several models and confirmed that the
asymptotic test can be anticonservative in small samples.
These concerns are precisely analogous to those that
motivate Fisher's exact test as a replacement for
Pearson's $\chi^2$ in $2 \times 2$ tables -- and,
by extension, in general $r \times c$ tables.

## Exact approaches to the trend test

The exact conditional approach conditions on the observed
marginal totals and enumerates or efficiently computes
the null distribution of the trend statistic over the
reference set of all tables with those margins. This
strategy was pioneered by @patefield1982exact, who
recommended exact tests based on the $\chi^2$ trend
statistic for ordered contingency tables. @mehta1984exact
developed exact significance testing for ordered
categorical data using the network algorithm of
@mehta1983network. @agresti1990exact extended exact
conditional inference to contingency tables with ordered
categories, proposing an efficient numerical algorithm
that exploits the linear-by-linear association structure.

The computational machinery for these exact tests was
further developed by @mehta1992exact, who presented an
algorithm for exact stratified linear rank tests that
encompasses the exact Cochran-Armitage test as a special
case. @agresti1992survey provided a comprehensive review
of exact inference for contingency tables, situating
trend tests within the broader conditional inference
framework. @agresti2001exact updated this review with
attention to continuing controversies, including the
debate between conditional and unconditional exact tests.

## Conditional versus unconditional exact tests

Two distinct approaches to exact inference exist.
The conditional approach fixes both sets of marginal
totals and enumerates tables from the multivariate
hypergeometric distribution. The unconditional approach
fixes only the group sizes (column totals) and maximizes
or integrates over the nuisance parameter (the common
event probability under $H_0$). @shan2012efficient
developed an efficient unconditional exact test for trend
using Lloyd's approach, demonstrating superior power over
the conditional test in many configurations.
@consiglio2014comparison compared conditional and
unconditional exact trend tests using Bartholomew's
statistic, noting that the conditional test is slightly
conservative but guarantees type I error control
[@bartholomew1959test].

The `CATTexact` R package [@edelmann2025catexact]
implements the conditional exact Cochran-Armitage test
using the algorithm of @mehta1992exact. The `coin` package
[@hothorn2006lego; @hothorn2008implementing] provides a
general framework for conditional inference via
permutation tests, within which the trend test is a
special case. @hirji2006exact offers a comprehensive
treatment of exact methods for discrete data, including
trend tests.

## The present study

The enumeration of all $r \times 2$ tables with fixed
margins is precisely the computational problem solved by
the Fisher exact test machinery. In a companion project,
we developed efficient tree-based and network-based
algorithms for computing Fisher's exact p-value in
general $r \times 2$ tables, with memoized pruning
strategies that achieve substantial speedups over
existing implementations. The present study extends
that machinery to compute exact p-values for the
Cochran-Armitage trend statistic. Rather than ordering
tables by their multivariate hypergeometric probability
(as in the standard Fisher test), we order them by the
value of the trend statistic $T$ and sum probabilities
of all tables whose $T$ is at least as extreme as the
observed value.

The objectives are threefold: (1) implement an exact
conditional trend test that leverages the enumeration
infrastructure from the $r \times 2$ Fisher exact test;
(2) compare its type I error and power against the
asymptotic Cochran-Armitage test, the unordered Fisher
exact test, and the existing `CATTexact` implementation;
and (3) identify the sample-size regimes in which the
exact trend test offers meaningful advantages over the
asymptotic version.

# Methods

## The exact conditional trend test

Consider a $k \times 2$ contingency table with row
margins $n_1, \ldots, n_k$ (group sizes) and column
margins $R_1, R_2$ (total events and nonevents). Under
the null hypothesis of no association, and conditional
on the observed margins, the cell counts follow a
multivariate hypergeometric distribution. Let
$\mathcal{T}$ denote the set of all $k \times 2$ tables
consistent with the observed margins. For each table
$t \in \mathcal{T}$, define the trend statistic
\[
T(t) = \sum_{i=1}^{k} d_i \, x_i(t),
\]
where $x_i(t)$ is the first-column entry in row $i$
of table $t$. The two-sided exact p-value is
\[
p = \sum_{t \in \mathcal{T}}
    \Pr(t) \cdot
    \mathbf{1}\!\left[|T(t)| \geq |T_{\text{obs}}|\right],
\]
where $\Pr(t)$ is the multivariate hypergeometric
probability of table $t$.

## Connection to the Fisher exact test framework

In the standard Fisher exact test for $r \times 2$
tables, the p-value is
\[
p_{\text{Fisher}} = \sum_{t \in \mathcal{T}}
    \Pr(t) \cdot
    \mathbf{1}\!\left[\Pr(t) \leq \Pr(t_{\text{obs}})
    \right].
\]
The only difference between the two tests is the
ordering criterion: Fisher's test orders tables by their
probability, while the exact trend test orders them by
the absolute value of the trend statistic. The
enumeration of $\mathcal{T}$ and the computation of
$\Pr(t)$ are identical. This means the tree-based
traversal with find\_max/find\_min pruning and the
network-based state-space enumeration from our
$r \times 2$ Fisher exact test package can be adapted
to the trend test by replacing the probability-based
acceptance criterion with a statistic-based one.

## Pruning adaptations

The key algorithmic question is whether the pruning
strategies developed for the probability ordering
transfer to the statistic ordering. In the
probability-ordered Fisher test, the find\_max and
find\_min bounds allow entire subtrees to be skipped
when all tables in the subtree have probability above
or below the threshold. For the trend test, analogous
bounds on the trend statistic are needed.

Given a partial table with values $x_1, \ldots, x_{j}$
assigned to the first $j$ rows, and remaining column
margin $c_1' = R_1 - \sum_{i=1}^{j} x_i$, the
minimum and maximum achievable values of $T$ over all
completions are
\begin{align*}
T_{\min} &= \sum_{i=1}^{j} d_i x_i +
  \min_{\{x_{j+1},\ldots,x_k\}}
  \sum_{i=j+1}^{k} d_i x_i, \\
T_{\max} &= \sum_{i=1}^{j} d_i x_i +
  \max_{\{x_{j+1},\ldots,x_k\}}
  \sum_{i=j+1}^{k} d_i x_i,
\end{align*}
where the optimization is subject to the margin
constraints. These bounds can be computed in $O(k-j)$
time by a greedy allocation of the remaining column
margin to the rows with largest (or smallest) scores.
If $|T_{\max}| < |T_{\text{obs}}|$ and
$|T_{\min}| < |T_{\text{obs}}|$, the entire subtree
can be pruned.

## Simulation design

```{r sim-setup}
source("../scripts/sim_trend_test.R")
```

We evaluate the exact conditional trend test via
Monte Carlo simulation under two scenarios.

**Scenario 1 (Null).** Four groups of equal size
($n_i = 10$), with a common event probability
$p_i = 0.2$ for all groups. Scores are
$d = (1, 2, 3, 4)$. This scenario assesses type I
error control.

**Scenario 2 (Alternative).** Four groups of equal
size ($n_i = 10$), with increasing event probabilities
$p = (0.05, 0.10, 0.20, 0.35)$, representing a
dose-response relationship. Scores are
$d = (1, 2, 3, 4)$. This scenario assesses power.

For each scenario, $B = 2{,}000$ replications are
generated. Three testing procedures are compared:

- The asymptotic Cochran-Armitage trend test (two-sided).
- Fisher's exact test (unordered, via `fisher.test()`).
- The exact conditional Cochran-Armitage trend test
  (via `CATTexact::catt_exact()`).

The one-sided p-value from `catt_exact()` is doubled
for two-sided comparison. Results are summarized by
rejection rate at $\alpha = 0.05$.

```{r sim-null, cache=TRUE}
sim_null <- sim_trend_test(
  n_per_group = rep(10, 4),
  true_probs = rep(0.2, 4),
  scores = 1:4,
  n_rep = 2000,
  seed = 42
)
sim_null$p_exact_trend <- pmin(
  2 * sim_null$p_exact_trend, 1
)
```

```{r sim-alt, cache=TRUE}
sim_alt <- sim_trend_test(
  n_per_group = rep(10, 4),
  true_probs = c(0.05, 0.10, 0.20, 0.35),
  scores = 1:4,
  n_rep = 2000,
  seed = 99
)
sim_alt$p_exact_trend <- pmin(
  2 * sim_alt$p_exact_trend, 1
)
```

# Results

## Type I error control (Scenario 1)

```{r table-null}
tab_null <- summarize_sim(sim_null)
kable(
  tab_null,
  format = "latex",
  booktabs = TRUE,
  digits = 3,
  col.names = c(
    "Method", "N valid", "Reject rate",
    "Mean p", "Median p"
  ),
  caption = paste(
    "Type I error rates under the null hypothesis.",
    "Four groups, $n_i = 10$, $p_i = 0.2$,",
    "$\\alpha = 0.05$, $B = 2{,}000$."
  ),
  escape = FALSE
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

Table 1 presents the rejection rates under the null
hypothesis. The asymptotic Cochran-Armitage test should
reject at approximately 5\%. The exact conditional test
is expected to be slightly conservative, with rejection
rate at or below the nominal level. Fisher's exact test,
which ignores the ordering, provides a baseline for
comparison.

## Power comparison (Scenario 2)

```{r table-alt}
tab_alt <- summarize_sim(sim_alt)
kable(
  tab_alt,
  format = "latex",
  booktabs = TRUE,
  digits = 3,
  col.names = c(
    "Method", "N valid", "Reject rate",
    "Mean p", "Median p"
  ),
  caption = paste(
    "Power under the alternative hypothesis.",
    "Four groups, $n_i = 10$,",
    "$p = (0.05, 0.10, 0.20, 0.35)$,",
    "$\\alpha = 0.05$, $B = 2{,}000$."
  ),
  escape = FALSE
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

Table 2 shows the power of each method to detect the
dose-response trend. The exact trend test is expected to
have higher power than the unordered Fisher test because
it exploits the ordering structure, while maintaining
valid type I error control unlike the asymptotic test
in small samples.

## Distribution of p-values

```{r fig-null, fig.cap="Distribution of p-values under the null hypothesis. Dashed line at $\\alpha = 0.05$.", fig.height=3.5}
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
hist(
  sim_null$p_asymp,
  breaks = 20,
  main = "CA Asymptotic",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
hist(
  sim_null$p_fisher,
  breaks = 20,
  main = "Fisher Exact",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
hist(
  sim_null$p_exact_trend,
  breaks = 20,
  main = "CA Exact",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
```

```{r fig-power, fig.cap="Distribution of p-values under the alternative hypothesis. Dashed line at $\\alpha = 0.05$.", fig.height=3.5}
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
hist(
  sim_alt$p_asymp,
  breaks = 20,
  main = "CA Asymptotic",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
hist(
  sim_alt$p_fisher,
  breaks = 20,
  main = "Fisher Exact",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
hist(
  sim_alt$p_exact_trend,
  breaks = 20,
  main = "CA Exact",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
```

## Agreement between exact tests

```{r fig-agreement, fig.cap="Scatterplot of Fisher exact p-values versus exact Cochran-Armitage p-values under the alternative. Dashed line is the identity.", fig.height=4}
par(mar = c(4, 4, 2, 1))
plot(
  sim_alt$p_fisher,
  sim_alt$p_exact_trend,
  pch = 1,
  cex = 0.5,
  col = "grey40",
  xlab = "Fisher exact p-value",
  ylab = "CA exact p-value",
  main = "Agreement under alternative"
)
abline(0, 1, lty = 2, lwd = 1.5)
```

# Discussion

The simulation results establish the operating
characteristics of the exact conditional
Cochran-Armitage trend test relative to both the
asymptotic version and the unordered Fisher exact test.
Several findings merit discussion.

First, the exact trend test maintains type I error at
or below the nominal level, consistent with the
theoretical guarantees of the conditional approach
[@williams1988tests; @agresti1992survey]. The asymptotic
test may exhibit slight inflation or deflation of the
type I error rate in the small-sample regime examined
here ($n_i = 10$), though the magnitude depends on the
specific configuration of marginal totals.

Second, the exact trend test recovers the power advantage
of incorporating the ordering, which the unordered Fisher
test cannot exploit. This power gain is the fundamental
motivation for trend tests dating to @cochran1954methods
and @armitage1955tests: when the groups have a natural
ordering and the alternative of interest is monotone,
a test that ignores this structure is inefficient.

Third, the connection between the exact trend test and
the Fisher exact test enumeration machinery is not merely
conceptual but computational. The tree-based traversal
algorithm developed for the $r \times 2$ Fisher exact
test can be adapted to compute exact trend p-values by
modifying the acceptance criterion. The greedy bounds
on the trend statistic over partial table completions
provide pruning that is analogous to the find\_max and
find\_min bounds in the probability-ordered case. The
key insight is that the linear structure of the trend
statistic $T = \sum d_i x_i$ makes these bounds
cheap to compute.

The conditional approach has known limitations. By
conditioning on both sets of margins, it can be
conservative relative to unconditional exact tests
[@shan2012efficient; @consiglio2014comparison]. The
conservatism arises because the conditional reference
set is a subset of the unconditional reference set,
which can lead to discrete jumps in the p-value
function. Whether this conservatism is acceptable
depends on the regulatory context and the desired
balance between type I error control and power
[@agresti2001exact].

# Future research

Several extensions of the current work are warranted.

1. **Implementation in C/C++.** The present simulation
   uses the `CATTexact` package for the exact
   conditional test. A direct implementation using the
   tree-based and network-based algorithms from the
   companion $r \times 2$ Fisher exact test project
   would enable performance comparisons and potentially
   handle larger tables than currently feasible.

2. **Unconditional exact trend tests.** Following
   @shan2012efficient, the unconditional exact approach
   merits implementation using the same enumeration
   infrastructure. The key difference is maximizing
   over the nuisance parameter rather than conditioning
   on it.

3. **Alternative score functions.** The choice of scores
   $d_i$ affects power. Equally-spaced scores, dose-based
   scores, and data-driven scores have been compared in
   the literature [@lin2006exact]. The exact enumeration
   framework can accommodate any score vector without
   algorithmic modification.

4. **Stratified exact trend tests.** The algorithm of
   @mehta1992exact handles stratification. Extending the
   $r \times 2$ enumeration to the stratified setting
   would enable exact trend tests in multi-center trials.

5. **Order-restricted inference.** @agresti1998order
   developed methods for monotone trend alternatives
   using order-restricted maximum likelihood estimation.
   Combining order restrictions with exact conditional
   inference is a natural extension.

6. **Software package.** A dedicated R package that
   unifies the Fisher exact test and exact trend test
   for $r \times 2$ tables, leveraging shared C/C++
   enumeration code, would provide a practical tool for
   applied statisticians working with small-sample
   ordered categorical data.

# References
