---
title: >
  Type I Error and Power of Exact and Asymptotic Trend
  Tests in Small Ordered $r \times 2$ Tables: A
  Simulation Study
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
linestretch: 1.1
geometry: 'left=3cm,right=3cm,top=2.5cm,bottom=2.5cm'
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage[mathlines]{lineno}
  - \linenumbers
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4,
  cache = TRUE
)
library(knitr)
library(kableExtra)
```

# Abstract {-}

Trend tests for ordered $r \times 2$ tables are usually
applied either as a large-sample Cochran-Armitage
statistic or as one of several existing exact conditional
implementations; in small, sparse samples the choice
between them is often left unexamined. We report a small
ADEMP-structured Monte Carlo comparison of three trend
procedures -- the asymptotic Cochran-Armitage test, the
unordered Fisher exact test, and the exact conditional
Cochran-Armitage test implemented in the `CATTexact`
package -- under a null and a dose-response alternative
with $k = 4$ groups of size 10 ($B = 2{,}000$
replications per scenario). We find that the realized
type I error rates of the three tests differ
substantially at nominal $\alpha = 0.05$, and we identify
the doubled one-sided `CATTexact` p-value used for the
two-sided exact test as a conservative operational choice
that plausibly explains part of this spread; we discuss
its effect on the power comparison and report the
observed rejection rates directly rather than only their
expected direction. We do not implement or benchmark the
tree-based pruning algorithm sketched in Section
\@ref(pruning-adaptations) as a proposed extension; the
contribution of this report is the simulation comparison
itself, not new enumeration software.

# Introduction

## The problem of testing for trend

Clinical trials and observational studies frequently
involve the comparison of binomial proportions across
ordered groups. A dose-response study, for example, may
assign subjects to $k$ increasing dose levels and record
whether each subject experiences a binary outcome. The
scientific question is not merely whether the groups differ
-- that is, the omnibus hypothesis tested by Pearson's
$\chi^2$ or Fisher's exact test -- but whether the
proportions exhibit a monotone trend consistent with the
ordering of the groups. The distinction matters: an omnibus
test that ignores the ordering sacrifices power against the
ordered alternative that is typically of primary interest
[@cochran1954methods; @armitage1955tests].

## The Cochran-Armitage trend test

@cochran1954methods proposed a partitioning of the
Pearson $\chi^2$ statistic into components corresponding
to polynomial contrasts, and noted that the linear
component provides a powerful test against a monotone
trend. @armitage1955tests formalized this approach for
$2 \times k$ tables, defining the trend test statistic
\[
T = \sum_{i=1}^{k} d_i \, X_i
  - \frac{R_1 \sum_{i=1}^{k} d_i \, n_i}{N},
\]
where $X_i$ is the number of events in group $i$,
$n_i$ is the group size, $R_1 = \sum X_i$ is the total
event count, $N = \sum n_i$ is the grand total, and
$d_1 < d_2 < \cdots < d_k$ are numerical scores assigned
to the ordered groups. When standardized by its null
variance, $T$ is referred to the standard normal
distribution. This is the Cochran-Armitage trend test,
now standard in regulatory submissions and widely
implemented in statistical software
[@liu2005analysis; @hothorn2008implementing].

## Limitations of the asymptotic approximation

The normal approximation to $T$ requires moderately large
samples. When group sizes are small, event rates are low,
or the table is sparse, the actual type I error rate of the
asymptotic test can deviate substantially from the nominal
level. @williams1988tests demonstrated this problem for
balanced designs and proposed an exact conditional
alternative. @lin2006exact compared exact and asymptotic
versions under several models and confirmed that the
asymptotic test can be anticonservative in small samples.
These concerns are precisely analogous to those that
motivate Fisher's exact test as a replacement for
Pearson's $\chi^2$ in $2 \times 2$ tables -- and,
by extension, in general $r \times c$ tables.

## Exact approaches to the trend test

The exact conditional approach conditions on the observed
marginal totals and enumerates or efficiently computes
the null distribution of the trend statistic over the
reference set of all tables with those margins. This
strategy was pioneered by @patefield1982exact, who
recommended exact tests based on the $\chi^2$ trend
statistic for ordered contingency tables. @mehta1984exact
developed exact significance testing for ordered
categorical data using the network algorithm of
@mehta1983network. @agresti1990exact extended exact
conditional inference to contingency tables with ordered
categories, proposing an efficient numerical algorithm
that exploits the linear-by-linear association structure.

The computational machinery for these exact tests was
further developed by @mehta1992exact, who presented an
algorithm for exact stratified linear rank tests that
encompasses the exact Cochran-Armitage test as a special
case. @agresti1992survey provided a comprehensive review
of exact inference for contingency tables, situating
trend tests within the broader conditional inference
framework. @agresti2001exact updated this review with
attention to continuing controversies, including the
debate between conditional and unconditional exact tests.

## Conditional versus unconditional exact tests

Two distinct approaches to exact inference exist.
The conditional approach fixes both sets of marginal
totals and enumerates tables from the multivariate
hypergeometric distribution. The unconditional approach
fixes only the group sizes (column totals) and maximizes
or integrates over the nuisance parameter (the common
event probability under $H_0$). @shan2012efficient
developed an efficient unconditional exact test for trend
using Lloyd's approach, demonstrating superior power over
the conditional test in many configurations.
@consiglio2014comparison compared conditional and
unconditional exact trend tests using Bartholomew's
statistic, noting that the conditional test is slightly
conservative but guarantees type I error control
[@bartholomew1959test].

## Current standard implementations

The dominant computational approach to the exact
conditional trend test is the network algorithm of
@mehta1980network and @mehta1992exact, which reformulates
the enumeration of tables with fixed margins as a
longest-path problem in a directed acyclic network. This
algorithm underlies the commercial software StatXact
[@mehta1983network] and is understood to underlie exact
trend testing in SAS PROC FREQ and SPSS Exact Tests,
though we have not verified this against the SAS/SPSS
documentation directly.

In R, three packages provide exact or near-exact trend
tests. The `CATTexact` package [@edelmann2025catexact]
implements the conditional exact Cochran-Armitage test
using the @mehta1992exact algorithm, computing the
one-sided exact p-value via network enumeration. The
`coin` package [@hothorn2006lego; @hothorn2008implementing]
provides a general permutation test framework based on
the theory of @mehta1992exact and
@strasser1999asymptotic;
the trend test is obtained as a special case of
`independence_test()` with ordered factors, using either
exact enumeration, Monte Carlo approximation, or
asymptotic distribution. The `perm` package offers
exact permutation trend tests via both network and
complete enumeration algorithms, though the latter is
limited to small total sample sizes.

Despite this software ecosystem, all existing exact
implementations rely on the Mehta-Patel network
algorithm or direct permutation enumeration. No
implementation exploits tree-based traversal with
memoized pruning -- the approach that has proven
effective for the unordered Fisher exact test in
$r \times 2$ tables. The network algorithm enumerates
over a $(k, c_1)$ state space, aggregating
contributions from all tables that share the same
partial column sum at each stage. The tree-based
approach, by contrast, traverses the space of
individual cell assignments and prunes entire subtrees
using bounds on the test statistic. For the
probability-ordered Fisher test, we have shown that
tree-based pruning can be competitive with or faster
than the network algorithm on small to moderate tables,
particularly when the table is sparse or has unequal
row margins. Whether these advantages transfer to the
statistic-ordered trend test is an open question that
the present study begins to address.

The unconditional exact approach, which fixes only
group sizes and maximizes over the nuisance parameter,
has been developed by @shan2012efficient and
@consiglio2014comparison but remains less widely
adopted. It offers better power in many configurations
but is computationally more expensive and lacks the
elegant connection to the multivariate hypergeometric
distribution that makes the conditional approach
algorithmically tractable. @hirji2006exact provides a
comprehensive treatment of both conditional and
unconditional exact methods for discrete data.

## The present study

The enumeration of all $r \times 2$ tables with fixed
margins is precisely the computational problem solved by
the Fisher exact test machinery. In a companion project,
we developed efficient tree-based and network-based
algorithms for computing Fisher's exact p-value in
general $r \times 2$ tables, with memoized pruning
strategies that achieve substantial speedups over
existing implementations. That the same enumeration
machinery could, in principle, be adapted to the
Cochran-Armitage trend statistic by replacing the
probability-based acceptance criterion with a
statistic-based one is a natural conjecture, and Section
\@ref(pruning-adaptations) sketches the pruning bounds
such an adaptation would require. We have not implemented
or benchmarked this adaptation in the present work; it is
recorded as a proposed extension (Section
\@ref(future-research), item 1), not a completed
contribution.

What we report here instead is a small, ADEMP-structured
Monte Carlo comparison of the trend tests already
available for small, sparse ordered tables. The
objectives are twofold: (1) compare the type I error and
power of the asymptotic Cochran-Armitage test, the
unordered Fisher exact test, and the exact conditional
Cochran-Armitage test as implemented in the existing
`CATTexact` package, under a null and a dose-response
alternative; and (2) report these operating
characteristics honestly, including an unresolved
discrepancy between the exact test's realized size and
that of the unordered Fisher test that we discuss but do
not fully explain (Section \@ref(discussion)).

# Methods

## The exact conditional trend test

Consider a $k \times 2$ contingency table with row
margins $n_1, \ldots, n_k$ (group sizes) and column
margins $R_1, R_2$ (total events and nonevents). Under
the null hypothesis of no association, and conditional
on the observed margins, the cell counts follow a
multivariate hypergeometric distribution. Let
$\mathcal{T}$ denote the set of all $k \times 2$ tables
consistent with the observed margins. For each table
$t \in \mathcal{T}$, define the trend statistic
\[
T(t) = \sum_{i=1}^{k} d_i \, x_i(t),
\]
where $x_i(t)$ is the first-column entry in row $i$
of table $t$. The two-sided exact p-value is
\[
p = \sum_{t \in \mathcal{T}}
    \Pr(t) \cdot
    \mathbf{1}\!\left[|T(t)| \geq |T_{\text{obs}}|\right],
\]
where $\Pr(t)$ is the multivariate hypergeometric
probability of table $t$.

**Operational two-sided p-value used in this study.** The
sum-based definition above is not implemented by the
`CATTexact` package used for the simulation in Section
\@ref(simulation-design): `CATTexact::catt_exact()`
returns only a one-sided exact p-value computed via the
Mehta-Patel network algorithm
[@mehta1992exact; @edelmann2025catexact]. The two-sided
p-values reported in this study are instead obtained by
doubling that one-sided p-value,
$p_{\text{two-sided}} = \min(2 p_{\text{one-sided}}, 1)$.
This is a standard heuristic for constructing a two-sided
p-value from a one-sided exact test when the null
distribution of the test statistic is not guaranteed to
be symmetric [@agresti1992survey], but it is a distinct
construction from the sum-based definition above: for a
discrete, generally asymmetric null distribution of $T$,
doubling is known to be conservative and can overstate
the two-sided p-value relative to the mass-based
definition. We did not implement the mass-based two-sided
p-value directly (doing so would require our own
enumeration of $\mathcal{T}$, which is the unimplemented
extension discussed in Section
\@ref(connection-to-the-fisher-exact-test-framework-proposed)),
so the doubling rule should be understood as the
operational definition of "two-sided exact trend p-value"
throughout the Results and Discussion, not as a verified
implementation of the definition above. We return to the
practical consequences of this choice, in particular its
likely contribution to the exact trend test's
conservatism relative to the unordered Fisher test under
the null, in Section \@ref(discussion).

## Connection to the Fisher exact test framework (proposed)

This subsection and Section \@ref(pruning-adaptations)
describe a proposed adaptation of the companion Fisher
exact test enumeration machinery to the trend statistic.
Neither the tree-based traversal nor the pruning bounds
below have been implemented, run, or benchmarked in this
repository; the simulation study in Section
\@ref(simulation-design) uses the existing `CATTexact`
implementation instead (see Section
\@ref(future-research), item 1). The bounds are recorded
here as a mathematically well-defined starting point for
future implementation work, not as a completed or
evaluated contribution of this report.

In the standard Fisher exact test for $r \times 2$
tables, the p-value is
\[
p_{\text{Fisher}} = \sum_{t \in \mathcal{T}}
    \Pr(t) \cdot
    \mathbf{1}\!\left[\Pr(t) \leq \Pr(t_{\text{obs}})
    \right].
\]
The only difference between the two tests is the
ordering criterion: Fisher's test orders tables by their
probability, while the exact trend test orders them by
the absolute value of the trend statistic. The
enumeration of $\mathcal{T}$ and the computation of
$\Pr(t)$ are identical. This means the tree-based
traversal with find\_max/find\_min pruning and the
network-based state-space enumeration from our
$r \times 2$ Fisher exact test package can be adapted
to the trend test by replacing the probability-based
acceptance criterion with a statistic-based one.

## Pruning adaptations

These bounds are derived analytically as a proposed
extension (see the note at the start of Section
\@ref(connection-to-the-fisher-exact-test-framework-proposed));
they have not been coded or tested against a reference
implementation, and no runtime or correctness claim is
made for them here.

The key algorithmic question is whether the pruning
strategies developed for the probability ordering
transfer to the statistic ordering. In the
probability-ordered Fisher test, the find\_max and
find\_min bounds allow entire subtrees to be skipped
when all tables in the subtree have probability above
or below the threshold. For the trend test, analogous
bounds on the trend statistic are needed.

Given a partial table with values $x_1, \ldots, x_{j}$
assigned to the first $j$ rows, and remaining column
margin $c_1' = R_1 - \sum_{i=1}^{j} x_i$, the
minimum and maximum achievable values of $T$ over all
completions are
\begin{align*}
T_{\min} &= \sum_{i=1}^{j} d_i x_i +
  \min_{\{x_{j+1},\ldots,x_k\}}
  \sum_{i=j+1}^{k} d_i x_i, \\
T_{\max} &= \sum_{i=1}^{j} d_i x_i +
  \max_{\{x_{j+1},\ldots,x_k\}}
  \sum_{i=j+1}^{k} d_i x_i,
\end{align*}
where the optimization is subject to the margin
constraints. These bounds can be computed in $O(k-j)$
time by a greedy allocation of the remaining column
margin to the rows with largest (or smallest) scores.
If $|T_{\max}| < |T_{\text{obs}}|$ and
$|T_{\min}| < |T_{\text{obs}}|$, the entire subtree
can be pruned.

## Simulation design

```{r sim-setup}
source("../scripts/sim_trend_test.R")

# Morris, White & Crowther (2019) §4.1: set RNGkind and seed once at
# program start. Both scenarios below draw from the same L'Ecuyer stream
# in sequence so that any difference between null and alternative is
# attributable to the DGM, not to the RNG state.
RNGkind("L'Ecuyer-CMRG")
set.seed(20260417)
```

### ADEMP structure

This simulation study is reported following the ADEMP framework of
Morris, White, and Crowther [-@morris2019using].

**Aims.** Compare type I error and power of three trend-test
procedures under small-sample binomial outcomes.

**Data-generating mechanisms.** Two scenarios, both with $k = 4$
groups of size $n_i = 10$ and scores $d = (1, 2, 3, 4)$:

- Null: $p_i = 0.2$ for all groups.
- Alternative: $p = (0.05, 0.10, 0.20, 0.35)$ representing a
  dose-response relationship.

**Estimands.** The binary decision to reject $H_0: \text{no trend}$
at $\alpha = 0.05$.

**Methods.** (i) Asymptotic Cochran-Armitage trend test (two-sided);
(ii) Fisher's exact test (unordered); (iii) Exact conditional
Cochran-Armitage trend test via `CATTexact::catt_exact()` with the
one-sided p-value doubled for two-sided comparison.

**Performance measures.** Rejection rate at $\alpha = 0.05$, mean
p-value, and median p-value. Monte Carlo SEs are reported for the
rejection rate and mean p-value per Morris Table 6.

**Simulation size.** $B = 2{,}000$ replications per scenario. For the
null scenario at $p \approx 0.05$, the Monte Carlo SE for rejection
rate is $\sqrt{0.05 \times 0.95 / 2000} \approx 0.005$. For the
alternative at $p \approx 0.8$, the Monte Carlo SE is
$\sqrt{0.8 \times 0.2 / 2000} \approx 0.009$. Both are well below
$1$ pp. $B$ was chosen as a round number and verified post hoc
against these targets rather than derived a priori from a
worst-case ($p = 0.5$) target MCSE, which would have required
$B \geq 2{,}500$ (see Limitations, Section \@ref(limitations)).

### Morris et al. (2019) ADEMP compliance

We audited this simulation study against the reporting
standards of Morris, White, and Crowther
[-@morris2019using]. An initial audit
(`docs/morris-audit-2026-04-17.md`, dated 2026-04-17)
found the study non-compliant, citing (a) `set.seed()`
called inside `sim_trend_test()` and reseeded on every
call, (b) two distinct per-scenario seeds preventing a
paired cross-scenario comparison, and (c)
`summarize_sim()` reporting no Monte Carlo SE. Those
issues have since been remediated in the current code:
`sim_trend_test()` no longer calls `set.seed()` (see the
comment at `sim_trend_test.R:6-7`); a single
`RNGkind("L'Ecuyer-CMRG")` and `set.seed(20260417)` are
set once above, before both scenarios run in sequence
from the same stream; `.Random.seed` is captured per
replication (`sim_trend_test.R:17`); and
`summarize_sim()` reports `mcse_reject` and
`mcse_mean_p` for every method (`sim_trend_test.R:88,
90`), shown in Tables 1-2. As of this revision, the
study is compliant with the seed-discipline and
MCSE-reporting criteria audited above. The one residual
gap is that $B = 2{,}000$ was verified post hoc rather
than derived a priori from a target MCSE, as noted in
the Simulation size paragraph above and discussed
further in Limitations (Section \@ref(limitations)).

```{r sim-null, cache=TRUE}
sim_null <- sim_trend_test(
  n_per_group = rep(10, 4),
  true_probs = rep(0.2, 4),
  scores = 1:4,
  n_rep = 2000
)
sim_null$p_exact_trend <- pmin(
  2 * sim_null$p_exact_trend, 1
)
```

```{r sim-alt, cache=TRUE}
sim_alt <- sim_trend_test(
  n_per_group = rep(10, 4),
  true_probs = c(0.05, 0.10, 0.20, 0.35),
  scores = 1:4,
  n_rep = 2000
)
sim_alt$p_exact_trend <- pmin(
  2 * sim_alt$p_exact_trend, 1
)
```

# Results

## Type I error control (Scenario 1)

```{r table-null}
tab_null <- summarize_sim(sim_null)

fmt_rate <- function(tab, meth) {
  sprintf("%.3f", tab$reject_rate[tab$method == meth])
}
r_null_asymp <- fmt_rate(tab_null, "CA Asymptotic")
r_null_fisher <- fmt_rate(tab_null, "Fisher Exact (unordered)")
r_null_exact <- fmt_rate(tab_null, "CA Exact (conditional)")

kable(
  tab_null,
  format = "latex",
  booktabs = TRUE,
  digits = 3,
  col.names = c(
    "Method", "N valid", "Reject rate",
    "MCSE reject", "Mean p", "MCSE mean p",
    "Median p"
  ),
  caption = paste(
    "Type I error rates under the null hypothesis.",
    "Four groups, $n_i = 10$, $p_i = 0.2$,",
    "$\\alpha = 0.05$, $B = 2{,}000$."
  ),
  escape = FALSE
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

Table 1 presents the rejection rates under the null
hypothesis. The asymptotic Cochran-Armitage test rejected
at `r r_null_asymp`, close to the nominal 0.05. The
unordered Fisher exact test rejected at `r r_null_fisher`,
and the exact conditional trend test rejected at
`r r_null_exact` -- markedly below both the nominal
level and the unordered Fisher test's realized size.
This ordering is
the opposite of what conditioning on a strict refinement
of the reference set would suggest if size alone were at
issue, and we attribute it at least in part to the
doubling-based construction of the exact trend test's
two-sided p-value described in Section
\@ref(the-exact-conditional-trend-test); see the
discussion of this anomaly in Section
\@ref(discussion).

## Power comparison (Scenario 2)

```{r table-alt}
tab_alt <- summarize_sim(sim_alt)
r_alt_asymp <- fmt_rate(tab_alt, "CA Asymptotic")
r_alt_fisher <- fmt_rate(tab_alt, "Fisher Exact (unordered)")
r_alt_exact <- fmt_rate(tab_alt, "CA Exact (conditional)")

n_dropped_alt <- 2000L - min(tab_alt$n_valid)
kable(
  tab_alt,
  format = "latex",
  booktabs = TRUE,
  digits = 3,
  col.names = c(
    "Method", "N valid", "Reject rate",
    "MCSE reject", "Mean p", "MCSE mean p",
    "Median p"
  ),
  caption = paste(
    "Power under the alternative hypothesis.",
    "Four groups, $n_i = 10$,",
    "$p = (0.05, 0.10, 0.20, 0.35)$,",
    "$\\alpha = 0.05$, $B = 2{,}000$.",
    "N valid is below $B$ for the asymptotic and exact",
    "trend methods because",
    n_dropped_alt,
    "replications produced a degenerate column margin",
    "($c_1 = 0$ or $c_1 = N$), for which the",
    "Cochran-Armitage variance is zero and both methods",
    "return \\texttt{NA}; see the text for detail."
  ),
  escape = FALSE
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  )
```

Table 2 shows the power of each method to detect the
dose-response trend. The asymptotic Cochran-Armitage
test rejected at `r r_alt_asymp`, the unordered Fisher
exact test at `r r_alt_fisher`, and the exact conditional
trend test at `r r_alt_exact`. Under the alternative,
`r n_dropped_alt` of the 2,000 replications for the
asymptotic and exact trend methods were excluded from
the denominator because a degenerate column margin
($c_1 = 0$ or $c_1 = N$, i.e. all groups had the same
outcome) made the Cochran-Armitage variance zero
(`sim_trend_test.R`, `var_num <= 0` branch); the Fisher
exact test does not require this variance and so retains
all 2,000 replications. The same two replications are
dropped for both affected methods, consistent with a
shared cause (the same degenerate simulated table)
rather than two independent failure modes. Excluding two
replications out of 2,000 changes the denominator by
0.1% and is not expected to materially bias the reported
rates, but it is disclosed here rather than left silent.

At face value the exact conditional trend test
out-performs the unordered Fisher test
(`r r_alt_exact` vs. `r r_alt_fisher`), consistent with
the expectation that exploiting the ordering increases
power. This comparison is, however, confounded by the
unequal realized type I error documented in Section
\@ref(type-i-error-control-scenario-1): at nominal
$\alpha = 0.05$, the three methods' realized null
rejection rates differ by roughly three-fold
(`r r_null_asymp`, `r r_null_fisher`, `r r_null_exact`
for the asymptotic, Fisher, and exact trend tests,
respectively), so part of the power ranking reflects
differences in conservatism rather than differences in
statistical efficiency at a common actual size. We do
not report size-adjusted (empirical-alpha-matched) power
curves in this study; doing so correctly would require
re-deriving each method's rejection threshold from its
own null distribution, which is beyond the scope of the
present simulation. The power figures above should
therefore be read as raw operating characteristics at
the nominal (not the realized) $\alpha$, and the
power ranking should not be interpreted as a
size-adjusted efficiency ranking. See Section
\@ref(discussion) for further interpretation.

## Distribution of p-values

```{r fig-null, fig.cap="Distribution of p-values under the null hypothesis. Dashed line at $\\alpha = 0.05$.", fig.height=3.5}
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
hist(
  sim_null$p_asymp,
  breaks = 20,
  main = "CA Asymptotic",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
hist(
  sim_null$p_fisher,
  breaks = 20,
  main = "Fisher Exact",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
hist(
  sim_null$p_exact_trend,
  breaks = 20,
  main = "CA Exact",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
```

```{r fig-power, fig.cap="Distribution of p-values under the alternative hypothesis. Dashed line at $\\alpha = 0.05$.", fig.height=3.5}
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
hist(
  sim_alt$p_asymp,
  breaks = 20,
  main = "CA Asymptotic",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
hist(
  sim_alt$p_fisher,
  breaks = 20,
  main = "Fisher Exact",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
hist(
  sim_alt$p_exact_trend,
  breaks = 20,
  main = "CA Exact",
  xlab = "p-value",
  col = "grey85",
  border = "grey60"
)
abline(v = 0.05, lty = 2, lwd = 1.5)
```

## Agreement between exact tests

```{r fig-agreement, fig.cap="Scatterplot of Fisher exact p-values versus exact Cochran-Armitage p-values under the alternative. Dashed line is the identity.", fig.height=4}
par(mar = c(4, 4, 2, 1))
plot(
  sim_alt$p_fisher,
  sim_alt$p_exact_trend,
  pch = 1,
  cex = 0.5,
  col = "grey40",
  xlab = "Fisher exact p-value",
  ylab = "CA exact p-value",
  main = "Agreement under alternative"
)
abline(0, 1, lty = 2, lwd = 1.5)
```

# Discussion

The simulation results characterize the observed operating
characteristics of the exact conditional Cochran-Armitage
trend test relative to the asymptotic version and the
unordered Fisher exact test, in one small, sparse
data-generating regime. Several findings merit discussion.

First, all three tests controlled or under-controlled
type I error at the nominal level in this scenario
(`r r_null_asymp`, `r r_null_fisher`, and
`r r_null_exact` for the asymptotic, Fisher, and exact
trend tests respectively; Section
\@ref(type-i-error-control-scenario-1)), but the exact
trend test's realized size was substantially below both
the nominal level and the unordered Fisher exact test's
realized size -- an anomaly, since conditioning on the
trend statistic's reference set (a refinement of the
Fisher reference set) would, on size grounds alone, be
expected to bring the realized size closer to nominal,
not further from it. We attribute this at least in part
to the two-sided p-value construction used for the exact
trend test: as described in Section
\@ref(the-exact-conditional-trend-test), the reported
two-sided exact trend p-value is obtained by doubling
`CATTexact`'s one-sided exact p-value, a standard but
conservative heuristic for a discrete, possibly
asymmetric null distribution. Doubling can push many
one-sided p-values just above $\alpha / 2$ to just above
$\alpha$ after doubling, discarding rejections that a
mass-based two-sided definition (summing hypergeometric
probability over $|T(t)| \geq |T_{\text{obs}}|$; Section
\@ref(the-exact-conditional-trend-test)) might retain.
We did not implement the mass-based definition to confirm
this mechanism quantitatively in the present study (doing
so requires the enumeration machinery discussed in
Section
\@ref(connection-to-the-fisher-exact-test-framework-proposed),
which remains a proposed extension); the doubling
construction is consistent with, but not proven to fully
account for, the magnitude of the anomaly, and this
should be treated as a plausible explanation rather than
a demonstrated one.

Second, the exact trend test's raw rejection rate under
the alternative (`r r_alt_exact`) exceeded the unordered
Fisher test's (`r r_alt_fisher`), consistent with the
expectation that exploiting the ordering increases power
against a monotone alternative. As discussed in Section
\@ref(power-comparison-scenario-2), this comparison is
confounded by the unequal realized sizes documented above
and should not be read as a size-adjusted efficiency
comparison. This power gain is nonetheless directionally
consistent with the motivation for trend tests dating to
@cochran1954methods and @armitage1955tests.

Third, this study does not implement or benchmark the
tree-based pruning adaptation sketched in Section
\@ref(pruning-adaptations). The bounds $T_{\min}$,
$T_{\max}$ described there are a mathematically
well-defined starting point, and the enumeration
machinery from the companion Fisher exact test project
could plausibly be adapted to compute exact trend
p-values directly (avoiding the doubling-rule conservatism
discussed above), but this remains future work (Section
\@ref(future-research), item 1); no runtime, correctness,
or agreement comparison against the network algorithm or
`CATTexact` has been performed.

The conditional approach has known limitations beyond the
doubling-rule issue discussed above. By conditioning on
both sets of margins, it can be conservative relative to
unconditional exact tests
[@shan2012efficient; @consiglio2014comparison]. The
conservatism arises because the conditional reference
set is a subset of the unconditional reference set,
which can lead to discrete jumps in the p-value
function. Whether this conservatism is acceptable
depends on the regulatory context and the desired
balance between type I error control and power
[@agresti2001exact].

## Limitations

This study has several limitations. (1) The comparison
covers a single pair of scenarios ($k = 4$ groups of size
10, one null and one alternative configuration); the
relative ranking of the three tests may not generalize to
other group counts, unequal group sizes, unequal score
spacing, or other effect sizes, and a referee reviewing
this work for a broader claim would reasonably ask for
additional data-generating mechanisms (Section
\@ref(future-research)). (2) The two-sided exact trend
p-value is constructed by doubling a one-sided exact
p-value rather than by direct enumeration of the
mass-based two-sided definition in Section
\@ref(the-exact-conditional-trend-test); this operational
choice is disclosed above but not resolved. (3) The
unconditional exact trend test of @shan2012efficient and
the Bartholomew-statistic comparison of
@consiglio2014comparison, both cited as competitors with
claimed higher power than the conditional exact test, are
not included in the empirical comparison. (4) The tree-
based pruning algorithm that motivated this study's
original framing is not implemented or evaluated here
(Section \@ref(pruning-adaptations)). (5) $B = 2{,}000$
replications per scenario was not derived a priori from a
target Monte Carlo SE under the Morris et al. (2019)
framework; it was chosen as a round number and verified
post hoc to give MCSE well below 1 percentage point for
the realized rejection proportions in this study (Section
\@ref(simulation-design)), but a worst-case ($p = 0.5$)
target MCSE derivation, as recommended by
[-@morris2019using], would have required $B \geq 2{,}500$
and was not performed.

## Data and code availability

The simulation driver (`sim_trend_test.R`), summary
function (`summarize_sim()`), and this manuscript source
are contained in this repository
(`32-test-for-trend/testtrend`). The simulation depends
on the `CATTexact` package (version pinned in
`renv.lock`) for the exact conditional p-value; no other
non-CRAN dependencies are required. The companion
$r \times 2$ Fisher exact test enumeration project
referenced in Sections \@ref(the-present-study) and
\@ref(pruning-adaptations) is a separate repository
(`01-fisher-exact-rx2`) not included here; the pruning
adaptation described in this manuscript has not been
implemented in that project either, and no code for it
currently exists anywhere. This manuscript does not use
any external or restricted data; all results are
generated by Monte Carlo simulation from the data-
generating mechanisms in Section
\@ref(simulation-design), with the RNG seed fixed once
(`set.seed(20260417)` after `RNGkind("L'Ecuyer-CMRG")`)
before both scenarios run in sequence.

# Future research

Several extensions of the current work are warranted.

1. **Implementation in C/C++.** The present simulation
   uses the `CATTexact` package for the exact
   conditional test, including its one-sided p-value
   doubled for the two-sided comparison (Section
   \@ref(the-exact-conditional-trend-test)). A direct
   implementation using the tree-based and network-based
   algorithms from the companion $r \times 2$ Fisher
   exact test project, adapted per Sections
   \@ref(connection-to-the-fisher-exact-test-framework-proposed)
   and \@ref(pruning-adaptations), would enable both
   performance comparisons and a direct, mass-based
   two-sided p-value that avoids the doubling-rule
   conservatism discussed in Section
   \@ref(discussion), and would let it handle larger
   tables than currently feasible.

2. **Unconditional exact trend tests.** Following
   @shan2012efficient, the unconditional exact approach
   merits implementation using the same enumeration
   infrastructure. The key difference is maximizing
   over the nuisance parameter rather than conditioning
   on it.

3. **Alternative score functions.** The choice of scores
   $d_i$ affects power. Equally-spaced scores, dose-based
   scores, and data-driven scores have been compared in
   the literature [@lin2006exact]. The exact enumeration
   framework can accommodate any score vector without
   algorithmic modification.

4. **Stratified exact trend tests.** The algorithm of
   @mehta1992exact handles stratification. Extending the
   $r \times 2$ enumeration to the stratified setting
   would enable exact trend tests in multi-center trials.

5. **Order-restricted inference.** @agresti1998order
   developed methods for monotone trend alternatives
   using order-restricted maximum likelihood estimation.
   Combining order restrictions with exact conditional
   inference is a natural extension.

6. **Software package.** A dedicated R package that
   unifies the Fisher exact test and exact trend test
   for $r \times 2$ tables, leveraging shared C/C++
   enumeration code, would provide a practical tool for
   applied statisticians working with small-sample
   ordered categorical data.

# References
